\documentclass[../main]{subfiles}

\begin{document}

\chapter*{Preface}
    Let's say it plainly - this book is for no one. For those who
already know C++ well, it is just a bunch of idiotic examples.
On the other hand, for those who don't know it yet, this random set of pages
is also a bunch of idiotic examples. I don't expect anyone to want to
read it, but there is a small chance that after some rough party
with your collegues in a dormitory, you might start exploring GitHub,
having nothing better to do at 2AM, and stumble upon it accidentaly.
So this is my explanation for you, my lost friend.\newline

    I mainly wrote it for myself as a collection of things that I learned -
either because I found them interesting or I couldn't remember them.
These sweet notes aren't mean to learn you C++; they don't
explain the basics or cover super advanced topics. They are just compilation
of well-known features but stored in one place. All the code and text
were written by me or taken from the books listed in the bibliography.
There is also a few real-life cases that I encountered in my daily work.\newline

    Anyway, if you decided to read this book I don't have much to say - good luck.
By the way, in case when you notice a mistake, silly code example or something
else (and I think you notice, there are plenty of them) I would be very
grateful when you leave the comment. This is one of the reasons I made
my notes public. For the rest of you, who have ended your journey with this
sentence - good choice.\newline

   The \LaTeX\vspace{1pt} code for this document is in the repository.
In the future I'm going to add chapters devoted to C++20 and C++23 and
fill the missing text parts in \textit{Concurrency} one.

\begin{flushright}
    \textit{Mikul}
\end{flushright}
\end{document}